---
title: "STA3043S Bayesian Analysis"
subtitle: "Assignment One Memorandum: Hierarchical Bayesian Modelling"
author: "Ryan Daye (DYXRYA001), Alex Meyer(MYRALE008), Jonathan Stocken (STCJON004)"
date: "6 September 2026"
output:
  pdf_document:
    latex_engine: pdflatex
    fig_caption: true
    keep_tex: true
    toc: false
    toc_depth: 1
geometry: a4paper, margin=23mm
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{caption}
  - \usepackage{microtype}
  - \captionsetup{font=small,labelfont=bf}
  - \captionsetup[figure]{justification=centering,singlelinecheck=false}
  - \captionsetup[table]{justification=centering,singlelinecheck=false}
  - \setlength{\emergencystretch}{3em}
  - \widowpenalty=10000
  - \clubpenalty=10000
  - \displaywidowpenalty=10000
  - \raggedbottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
load("Data/STA3043S_Assignment1_Data.RData")
mydata <- machineData[["DYXRYA001"]]
```

\noindent\textbf{Dataset:} This analysis uses the data assigned to student number
\texttt{DYXRYA001}.

```{r, echo=FALSE}
# Split observations by machine and calculate the summaries used below.
machine_levels <- sort(unique(mydata$Machine))
y_by_machine <- split(mydata$Y, mydata$Machine)
n_j <- vapply(y_by_machine, length, integer(1))
ybar_j <- vapply(y_by_machine, mean, numeric(1)) # Mean of each machine
sd_j <- vapply(y_by_machine, sd, numeric(1))
min_j <- vapply(y_by_machine, min, numeric(1))
max_j <- vapply(y_by_machine, max, numeric(1))

# Create a data frame of the above statistics so that I can plot a table
df_machines <- data.frame(
  Machine = machine_levels,
  'n_j' = n_j, 
  'Mean Deviation' = round(ybar_j, 3),
  'SD' = round(sd_j, 3),
  'Minimum' = round(min_j, 3),
  'Maximum' = round(max_j, 3),
  check.names = FALSE
)

```

# Question 1: Exploratory Analysis

```{r, echo=FALSE}
library(knitr)
library(kableExtra)

df_machines |>
  kable(
    format = "latex",
    booktabs = TRUE,
    align = c("c", "c", "c", "c", "c", "c"),
    caption = "Machine-level exploratory summaries (volumes in ml)."
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(
    0,
    bold = TRUE)

```

The table summarises the overall performance of the seven filling machines. There are 153 bottle measurements in total, with the number of observations ranging from 5 for Machine 1 to 50 for Machine 7. The sample mean deviations range from -1.529 ml for Machine 7, which underfills on average, to 4.640 ml for Machine 2, which overfills on average. This suggests that using only a single overall mean could hide important differences between the machines. The sample standard deviations range from 1.127 to 2.873 ml. Given the small and unequal sample sizes, these values are broadly consistent with the assumed standard deviation of $\sigma = 2$, although Machine 3 shows the greatest variability.

```{r, echo = FALSE, machine-boxplot, fig.cap="Observed deviations by machine. Boxes summarise the empirical distributions; points show individual bottles; the dashed line marks the 500 ml target (zero deviation).", fig.pos="H", fig.align="center", fig.width=7, fig.height=4.2, out.width="90%"}
par(mar = c(4.2, 4.8, 0.8, 0.5), las = 1)

boxplot(
  Y ~ factor(Machine),
  data = mydata,
  col = "lightblue",
  border = "navy",
  xlab = "Machine",
  ylab = "Deviation from 500 ml (ml)",
  outline = FALSE
)

set.seed(69)

stripchart(
  Y ~ factor(Machine),
  data = mydata,
  vertical = TRUE,
  method = "jitter",
  pch = 16,
  col = adjustcolor("steelblue", alpha.f = 0.5),
  cex = 0.7,
  add = TRUE
)

abline(
  h = 0,
  lty = 2,
  lwd = 1.2,
  col = "red"
)

```

Figure 1 supports the findings from Table 1. Machine 2 clearly tends to overfill, as all its observations are above the zero-deviation line. In contrast, Machines 6 and 7 tend to underfill, with their medians and most observations below zero. Machines 3, 4 and 5 have observations on both sides of the target, although Machine 3 shows noticeably more variation. Machine 1 also appears to underfill slightly, but this is based on only five observations. Overall, the plot shows clear differences in both the average deviation and variability of the machines.


# Question 2

### (a)

The observation model is

\[
Y_{ij}\mid\theta_j \sim N(\theta_j,\sigma^2),
\]

so we may write

\[
Y_{ij}=\theta_j+\varepsilon_{ij},
\qquad
\varepsilon_{ij}\overset{\mathrm{iid}}{\sim}N(0,\sigma^2).
\]

The sample mean for machine \(j\) is therefore

\[
\begin{aligned}
\overline{Y}_j
&=\frac{1}{n_j}\sum_{i=1}^{n_j}Y_{ij} \\
&=\frac{1}{n_j}\sum_{i=1}^{n_j}
  \left(\theta_j+\varepsilon_{ij}\right) \\
&=\theta_j+\overline{\varepsilon}_j,
\end{aligned}
\]

where

\[
\overline{\varepsilon}_j
=\frac{1}{n_j}\sum_{i=1}^{n_j}\varepsilon_{ij}.
\]

Because the errors are independent,

\[
\begin{aligned}
\operatorname{Var}(\overline{\varepsilon}_j)
&=\operatorname{Var}\left(
  \frac{1}{n_j}\sum_{i=1}^{n_j}\varepsilon_{ij}
  \right) \\
&=\frac{1}{n_j^2}
  \sum_{i=1}^{n_j}\operatorname{Var}(\varepsilon_{ij}) \\
&=\frac{1}{n_j^2}(n_j\sigma^2) \\
&=\frac{\sigma^2}{n_j}.
\end{aligned}
\]

Thus,

\[
\overline{Y}_j\mid\theta_j
\sim N\left(\theta_j,\frac{\sigma^2}{n_j}\right).
\]

The population model is

\[
\theta_j\mid\mu\sim N(\mu,\tau^2).
\]

Equivalently, we may write

\[
\theta_j=\mu+u_j,
\qquad
u_j\sim N(0,\tau^2).
\]

Consequently,

\[
\overline{Y}_j
=\mu+u_j+\overline{\varepsilon}_j.
\]

Conditional on \(\mu\), the random variables \(u_j\) and
\(\overline{\varepsilon}_j\) are independent. Therefore,

\[
\begin{aligned}
\operatorname{E}(\overline{Y}_j\mid\mu)
&=\mu,\\
\operatorname{Var}(\overline{Y}_j\mid\mu)
&=\operatorname{Var}(u_j)
  +\operatorname{Var}(\overline{\varepsilon}_j)\\
&=\tau^2+\frac{\sigma^2}{n_j}.
\end{aligned}
\]

Since the sum of independent normally distributed random variables is
normally distributed, it follows that

\[
\boxed{
\overline{Y}_j\mid\mu
\sim
N\left(
\mu,\,
\tau^2+\frac{\sigma^2}{n_j}
\right).
}
\]

The term \(\tau^2\) represents the between-machine variability caused by
the machine-specific parameters \(\theta_j\) varying around \(\mu\).
The term \(\sigma^2/n_j\) represents the within-machine sampling
variability in the sample mean based on \(n_j\) observations.

### (b)

Define

\[
V_j=\tau^2+\frac{\sigma^2}{n_j}.
\]

From part (a),

\[
\overline{Y}_j\mid\mu\sim N(\mu,V_j).
\]

Assuming conditional independence across the \(J\) machines, the
likelihood for \(\mu\) is

\[
\begin{aligned}
L(\mu;\mathbf{y})
&=\prod_{j=1}^{J}
  \frac{1}{\sqrt{2\pi V_j}}
  \exp\left\{
  -\frac{(\overline{y}_j-\mu)^2}{2V_j}
  \right\}\\
&\propto
  \exp\left\{
  -\frac{1}{2}
  \sum_{j=1}^{J}
  \frac{(\overline{y}_j-\mu)^2}{V_j}
  \right\}.
\end{aligned}
\]

Using the Laplace prior

\[
\pi(\mu)=\frac{c}{2}\exp\{-c|\mu|\},
\qquad c>0,
\]

Bayes' theorem gives

\[
\pi(\mu\mid\mathbf{y})
\propto
\exp\left\{
-\frac{1}{2}
\sum_{j=1}^{J}
\frac{(\overline{y}_j-\mu)^2}{V_j}
\right\}
\exp\{-c|\mu|\}.
\]

To complete the square, define

\[
A=\sum_{j=1}^{J}\frac{1}{V_j},
\qquad
B=\sum_{j=1}^{J}\frac{\overline{y}_j}{V_j},
\qquad
C=\sum_{j=1}^{J}\frac{\overline{y}_j^2}{V_j}.
\]

Then

\[
\begin{aligned}
\sum_{j=1}^{J}
\frac{(\overline{y}_j-\mu)^2}{V_j}
&=
\sum_{j=1}^{J}
\frac{\overline{y}_j^2-2\mu\overline{y}_j+\mu^2}{V_j}\\
&=A\mu^2-2B\mu+C\\
&=A\left(\mu-\frac{B}{A}\right)^2
  +C-\frac{B^2}{A}.
\end{aligned}
\]

Define

\[
m=\frac{B}{A}
=
\frac{
\displaystyle\sum_{j=1}^{J}\frac{\overline{y}_j}{V_j}
}{
\displaystyle\sum_{j=1}^{J}\frac{1}{V_j}
}
\]

and

\[
v=\frac{1}{A}
=
\left(
\sum_{j=1}^{J}\frac{1}{V_j}
\right)^{-1}.
\]

Therefore,

\[
\sum_{j=1}^{J}
\frac{(\overline{y}_j-\mu)^2}{V_j}
=
\frac{(\mu-m)^2}{v}
+C-\frac{B^2}{A}.
\]

The final two terms do not depend on \(\mu\), so they are absorbed into
the proportionality constant. Hence,

\[
\boxed{
\pi(\mu\mid\mathbf{y})
\propto
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
\exp\{-c|\mu|\}.
}
\]

# Question 3

### (a)

From the result of Question 2, the posterior density of \(\mu\) is proportional to

$$
\pi(\mu\mid\mathbf y)
\propto
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
\exp\{-c|\mu|\}.
$$

The first term is proportional to the density of a Normal distribution
with mean \(m\) and variance \(v\). Therefore, a natural candidate
distribution for Acceptance Rejection sampling is

$$
h(\mu)=N(m,v).
$$

The density of this is:

$$
h(\mu)
=
\frac{1}{\sqrt{2\pi v}}
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}.
$$

This is an appropriate candidate because it matches the Gaussian component
of the target posterior exactly. Due to the absoult value of \(\mu\),
The remaining factor satisfies:

$$
0 < \exp\{-c|\mu|\} \leq 1
$$

for all \(\mu\). Hence the target density is bounded by a constant multiple
of \(h(\mu)\), as required for Acceptance Rejection sampling. Furthermore,
the candidate has heavier tails than the target because the target contains
the additional factor \(\exp\{-c|\mu|\}\)

### (b)

From part (a), choose the candidate density

$$
h(\mu)
=
\frac{1}{\sqrt{2\pi v}}
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\},
$$

so that

$$
\mu^\ast \sim N(m,v).
$$

The unnormalised target posterior density is

$$
\pi^\ast(\mu\mid\mathbf y)
=
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
\exp\{-c|\mu|\}.
$$

For Acceptance Rejection sampling, we require a constant \(C\) such that

$$
\pi^\ast(\mu\mid\mathbf y)
\leq C h(\mu)
$$

for all \(\mu\). The smallest suitable value of \(C\) is

$$
C
=
\max_{\mu}
\frac{\pi^\ast(\mu\mid\mathbf y)}{h(\mu)}.
$$

Now,

$$
\begin{aligned}
\frac{\pi^\ast(\mu\mid\mathbf y)}{h(\mu)}
&=
\frac{
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
\exp\{-c|\mu|\}
}{
\frac{1}{\sqrt{2\pi v}}
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
}\\
&=
\sqrt{2\pi v}\exp\{-c|\mu|\}.
\end{aligned}
$$

Since

$$
\exp\{-c|\mu|\}
$$

is maximised when

$$
|\mu|=0,
$$

that is, when \(\mu=0\), its maximum value is

$$
\exp(0)=1.
$$

Therefore,

$$
C=\sqrt{2\pi v}.
$$

The probability of accepting a candidate value \(\mu\) is

$$
\gamma(\mu)
=
\frac{
\pi^\ast(\mu\mid\mathbf y)
}{
C h(\mu)
}.
$$

Substituting the expressions for \(\pi^\ast(\mu\mid\mathbf y)\),
\(C\), and \(h(\mu)\),

$$
\begin{aligned}
\gamma(\mu)
&=
\frac{
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
\exp\{-c|\mu|\}
}{
\sqrt{2\pi v}
\left[
\frac{1}{\sqrt{2\pi v}}
\exp\left\{
-\frac{(\mu-m)^2}{2v}
\right\}
\right]
}\\
&=
\exp\{-c|\mu|\}.
\end{aligned}
$$

Since \(c=0.25\),

$$
\gamma(\mu)
=
\exp\{-0.25|\mu|\}.
$$

### (c)

Using the candidate distribution from part (a),

$$
\mu^\ast \sim N(m,v),
$$

and the acceptance probability derived in part (b),

$$
\gamma(\mu^\ast)
=
\exp\{-c|\mu^\ast|\},
$$

The following function implements the sampler and returns the accepted draws
together with the empirical acceptance rate.

```{r accept reject, echo=TRUE}
ar <- function(n = 10000, m, v, c = 0.25) {
  samples <- numeric(n)
  accepted <- 0
  proposed <- 0
  
  while (accepted < n) {
    mu_candidate <- rnorm(
      1,
      mean = m,
      sd = sqrt(v)
    )
    
    proposed <- proposed + 1
    gamma <- exp(-c * abs(mu_candidate))
    u <- runif(1)
    if (u <= gamma) {
      accepted <- accepted + 1
      samples[accepted] <- mu_candidate
    }
  }
  acceptance_rate <- accepted / proposed
  
  return(
    list(
      samples = samples,
      acceptance_rate = acceptance_rate
    )
  )
}
```

```{r, echo=FALSE}
sigma2 <- 4
tau2 <- 2.25
c <- 0.25

V_j <- tau2 + sigma2 / n_j

m <- sum(ybar_j / V_j) / sum(1 / V_j)

v <- 1 / sum(1 / V_j)

set.seed(69)

ar_results <- ar(
  n = 5000,
  m = m,
  v = v,
  c = c
)

mu_draws <- ar_results$samples
acceptance_table <- data.frame(
  `Quantity` = "Empirical acceptance rate",
  `Estimate` = round(ar_results$acceptance_rate, 4),
  check.names = FALSE
)

acceptance_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c"),
    caption = "Acceptance--Rejection sampling performance."
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)
```

### (d)

#### (i)

Using the Acceptance Rejection samples
\(\mu^{(1)},\ldots,\mu^{(N)}\) from the posterior distribution, the
posterior mean is estimated by

$$
\widehat{E[\mu\mid\mathbf y]}
=
\frac{1}{N}
\sum_{s=1}^{N}\mu^{(s)},
$$

and the posterior variance is estimated by

$$
\widehat{\operatorname{Var}(\mu\mid\mathbf y)}
=
\frac{1}{N-1}
\sum_{s=1}^{N}
\left(
\mu^{(s)}-\bar{\mu}
\right)^2.
$$

```{r Mean and variance, echo=FALSE}
posterior_mean <- mean(mu_draws)
posterior_variance <- var(mu_draws)

posterior_summary <- data.frame(
  `Posterior Quantity` = c(
    "Posterior mean",
    "Posterior variance"
  ),
  `Estimate` = c(
    round(posterior_mean, 3),
    round(posterior_variance, 3)
  ),
  `Units` = c(
    "ml",
    "ml²"
  ),
  check.names = FALSE
)

posterior_summary |>
  kable(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c", "c"),
    caption = "Estimated posterior mean and variance of the population mean deviation."
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)
```


#### (ii)

A \(95\%\) posterior credible interval for \(\mu\) can be estimated using
the \(2.5\%\) and \(97.5\%\) quantiles of the simulated posterior draws.

```{r Credibility interval, echo=FALSE}
credible_interval <- quantile(
  mu_draws,
  probs = c(0.025, 0.975)
)

credible_table <- data.frame(
  `Interval` = "95\\% posterior credible interval",
  `Lower Bound` = unname(round(credible_interval[1], 3)),
  `Upper Bound` = unname(round(credible_interval[2], 3)),
  `Units` = "ml",
  check.names = FALSE
)

credible_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    row.names = FALSE,
    align = c("l", "c", "c", "c"),
    caption = "Estimated 95\\% posterior credible interval for $\\mu$.",
    escape = FALSE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)
```



#### (iii)

The posterior probability that the population of machines overfills on
average is estimated by the proportion of posterior draws for which
\(\mu>0\):

$$
P(\mu>0\mid\mathbf y)
\approx
\frac{1}{N}
\sum_{s=1}^{N}
I\left(\mu^{(s)}>0\right).
$$

Similarly:

$$
P(|\mu|<0.5\mid\mathbf y)
$$

is estimated by the proportion of posterior draws lying between
\(-0.5\) ml and \(0.5\) ml.

```{r positive near zero, echo=FALSE}
prob_positive <- mean(mu_draws > 0)

prob_near_zero <- mean(abs(mu_draws) < 0.5)

probability_table <- data.frame(
  `Posterior Probability` = c(
    "$P(\\mu > 0 \\mid \\mathbf{y})$",
    "$P(|\\mu| < 0.5 \\mid \\mathbf{y})$"
  ),
  `Estimate` = c(
    round(prob_positive, 4),
    round(prob_near_zero, 4)
  ),
  `Percentage` = c(
    paste0(round(100 * prob_positive, 1), "\\%"),
    paste0(round(100 * prob_near_zero, 1), "\\%")
  ),
  check.names = FALSE
)

probability_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    row.names = FALSE,
    align = c("l", "c", "c"),
    caption = "Estimated posterior probabilities.",
    escape = FALSE
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)
```



#### (iv)

The posterior distribution of \(\mu\) can be visualised using a kernel
density estimate of the Acceptance--Rejection samples.

```{r posterior-density, echo=FALSE, fig.cap="Estimated posterior density of the population mean deviation, $\\mu$.", fig.pos="H", fig.align="center", fig.width=7, fig.height=4.2}
posterior_density <- density(mu_draws)

plot(
  posterior_density,
  lwd = 2,
  xlab = expression(mu),
  ylab = "Posterior density",
  main = ""
)

abline(
  v = 0,
  lty = 2,
  lwd = 1.2
)
```

The posterior mean of \(\mu\) is 0.373 ml, suggesting that the population of filling machines has a slight tendency to overfill relative to the 500 ml target. This is supported by

$$
P(\mu > 0 \mid y) = 0.7404,
$$

so there is approximately a 74\% posterior probability that the average machine deviation is positive.

However, the 95\% posterior credible interval is

$$
(-0.762,\;1.500)\text{ ml},
$$

which contains zero. This indicates that there is still substantial uncertainty about whether the population of machines systematically overfills or underfills the containers.

Furthermore,

$$
P(|\mu| < 0.5 \mid y) = 0.5354,
$$

Thus, there is only a slight posterior preference for population mean deviations
within 0.5 ml of the target. Conversely, approximately 46.5\% of the posterior
probability lies outside this range, so a practically relevant calibration bias
remains plausible.

The posterior density is unimodal and concentrated slightly to the right of zero, which is consistent with the posterior mean and
the probability \(P(\mu > 0 \mid y)\). At the same time, the density still assigns appreciable probability to values near and
below zero.

Overall, the results provide mild evidence of factory-wide overfilling, but the
evidence is not decisive and substantial uncertainty remains about both the sign
and practical size of the population mean deviation. These results are relevant
because \(\mu\) represents the average deviation across the population of filling
machines, so inference about \(\mu\) indicates whether the manufacturing process
is systematically miscalibrated at the factory level rather than only for
individual machines.

# Question 4

### (a)

as shown in question 2a the likelihood here is,

\[
\overline{Y}_j \mid \theta_j \sim N\left(\theta_j, \frac{\sigma^2}{n_j}\right).
\]

the given prior is 

\[
\theta_j \mid \mu \sim N(\mu, \tau^2),
\]

using Bayes theorem to find the posterior for \(\theta_j\),

\[
\pi(\theta_j \mid \mu, \mathbf{y})
\propto
f(\overline{y}_j \mid \theta_j)\,\pi(\theta_j \mid \mu).
\]

equaling

\[
\pi(\theta_j \mid \mu, \mathbf{y})
\propto
\exp\left\{-\frac{n_j(\overline{y}_j-\theta_j)^2}{2\sigma^2}\right\}
\exp\left\{-\frac{(\theta_j-\mu)^2}{2\tau^2}\right\}.
\]

Consider the exponent of \(\pi(\theta_j \mid \mu, \mathbf{y})\):

\[
-\frac{1}{2}
\left[
\frac{n_j(\overline{y}_j-\theta_j)^2}{\sigma^2}
+\frac{(\theta_j-\mu)^2}{\tau^2}
\right].
\]

expand 

\[
\begin{aligned}
\frac{n_j(\overline{y}_j-\theta_j)^2}{\sigma^2}
&=\frac{n_j}{\sigma^2}
  \left(\overline{y}_j^2-2\theta_j\overline{y}_j+\theta_j^2\right),\\[4pt]
\frac{(\theta_j-\mu)^2}{\tau^2}
&=\frac{1}{\tau^2}
  \left(\theta_j^2-2\theta_j\mu+\mu^2\right).
\end{aligned}
\]

collect like terms 

\[
A_j=\frac{n_j}{\sigma^2}+\frac{1}{\tau^2},
\qquad
B_j=\frac{n_j\overline{y}_j}{\sigma^2}+\frac{\mu}{\tau^2},
\qquad
C_j=\frac{n_j\overline{y}_j^2}{\sigma^2}+\frac{\mu^2}{\tau^2}.
\]

then

\[
\frac{n_j(\overline{y}_j-\theta_j)^2}{\sigma^2}
+\frac{(\theta_j-\mu)^2}{\tau^2}
=A_j\theta_j^2-2B_j\theta_j+C_j.
\]

complete the square

\[
A_j\theta_j^2-2B_j\theta_j+C_j
=A_j\left(\theta_j-\frac{B_j}{A_j}\right)^2
+C_j-\frac{B_j^2}{A_j}.
\]

Define

\[
m_j=\frac{B_j}{A_j},
\qquad
v_j=\frac{1}{A_j}.
\]

Therefore,

\[
\frac{n_j(\overline{y}_j-\theta_j)^2}{\sigma^2}
+\frac{(\theta_j-\mu)^2}{\tau^2}
=\frac{(\theta_j-m_j)^2}{v_j}
+C_j-\frac{B_j^2}{A_j}.
\]

The final two terms not involving \(\theta_j\) are dropped into the proportionallity constant
and finally substituting gives. 

\[
\pi(\theta_j \mid \mu, \mathbf{y})
\propto
\exp\left\{-\frac{(\theta_j-m_j)^2}{2v_j}\right\}.
\]

Hence,

\[
\boxed{
\theta_j \mid \mu, \mathbf{y} \sim N(m_j, v_j).
}
\]

### (b)

Recall from part (a) that

\[
m_j=\frac{B_j}{A_j},
\qquad
A_j=\frac{n_j}{\sigma^2}+\frac{1}{\tau^2},
\qquad
B_j=\frac{n_j\overline{y}_j}{\sigma^2}+\frac{\mu}{\tau^2}.
\]

The prior mean is:  \(\mu\)\
The sample mean is: \(\overline{y}_j\)

Define the weights to be:

\[
w_j=\frac{\tau^2}{V_j},
\qquad
1-w_j=\frac{\sigma^2/n_j}{V_j},
\qquad
V_j=\tau^2+\frac{\sigma^2}{n_j}.
\]

The weighted average of the prior mean and the sample mean are

\[
\begin{aligned}
w_j\overline{y}_j+(1-w_j)\mu
&=\frac{\tau^2}{V_j}\overline{y}_j+\frac{\sigma^2/n_j}{V_j}\mu\\
&=\frac{\tau^2\overline{y}_j+(\sigma^2/n_j)\mu}{V_j}\\
&=\frac{n_j\tau^2\overline{y}_j+\sigma^2\mu}{n_j\tau^2+\sigma^2}\\
&=\frac{n_j\overline{y}_j/\sigma^2+\mu/\tau^2}{n_j/\sigma^2+1/\tau^2}\\
&=\frac{B_j}{A_j}\\
&=m_j.
\end{aligned}
\]

Hence,

\[
\boxed{
m_j=w_j\overline{y}_j+(1-w_j)\mu.
}
\]

Here we can clearly see shrinkage at play as the posterior mean of \(\theta_j\) is a weighted average of the prior mean and the sample mean. The posterior mean is pulled away from its sample mean towards its shared population mean with weight \(1-w_j\). This weighting makes intuitive sense as the lower amount of observations for the sample mean make it more unreliable and less valuable to estimating the posterior mean this is included in the weighting as a low \(n_j\) results in the sample mean being weighted less. The same is also true for a large \(n_j\) this allows us to use the data more effectively allowing the mean to change depending on the data of that machine. 

### (c)

From part (a),

$$
\theta_j \mid \mu,\mathbf{y}
\sim
N(m_j,v_j),
$$

where

$$
v_j
=
\left(
\frac{n_j}{\sigma^2}
+
\frac{1}{\tau^2}
\right)^{-1}
$$

and

$$
m_j
=
v_j
\left(
\frac{n_j\overline{y}_j}{\sigma^2}
+
\frac{\mu}{\tau^2}
\right).
$$

To obtain samples from the marginal posterior

$$
\pi(\theta_j\mid\mathbf{y})
=
\int
\pi(\theta_j\mid\mu,\mathbf{y})
\pi(\mu\mid\mathbf{y})\,d\mu,
$$

we first use the Acceptance--Rejection samples
\(\mu^{(1)},\ldots,\mu^{(N)}\) obtained in Question 3. For each sampled
value \(\mu^{(s)}\), we then generate

$$
\theta_j^{(s)}
\sim
\pi(\theta_j\mid\mu^{(s)},\mathbf{y}).
$$

Repeating this for every posterior draw of \(\mu\) produces samples from
the marginal posterior distribution of \(\theta_j\) for each machine.

```{r theta-sampling, echo=FALSE}
# mu_draws contains the 5000 posterior samples from the AR sampler
N <- length(mu_draws)
J <- length(ybar_j)

# Matrix to store the posterior draws for each theta_j
theta_draws <- matrix(
  NA,
  nrow = N,
  ncol = J
)

set.seed(70)

for (j in 1:J) {

  # Conditional posterior variance of theta_j
  v_theta_j <- 1 / (
    n_j[j] / sigma2 +
    1 / tau2
  )

  # Conditional posterior mean for each sampled value of mu
  m_theta_j <- v_theta_j * (
    n_j[j] * ybar_j[j] / sigma2 +
    mu_draws / tau2
  )

  # Draw theta_j | mu, y
  theta_draws[, j] <- rnorm(
    N,
    mean = m_theta_j,
    sd = sqrt(v_theta_j)
  )
}

colnames(theta_draws) <- paste0("Machine ", machine_levels)
```

The posterior mean of each \(\theta_j\) is estimated using the mean of its
simulated marginal posterior draws.

```{r theta-summary, echo=FALSE}
theta_posterior_means <- colMeans(theta_draws)

theta_intervals <- t(
  apply(
    theta_draws,
    2,
    quantile,
    probs = c(0.025, 0.975)
  )
)

theta_summary <- data.frame(
  `Machine` = machine_levels,
  `n_j` = n_j,
  `Observed Mean` = round(ybar_j, 3),
  `Posterior Mean` = round(theta_posterior_means, 3),
  `Lower 95%` = round(theta_intervals[, 1], 3),
  `Upper 95%` = round(theta_intervals[, 2], 3),
  check.names = FALSE
)

theta_summary |>
  kable(
    format = "latex",
    booktabs = TRUE,
    align = c("c", "c", "c", "c", "c", "c"),
    caption = "Observed and posterior machine-specific mean deviations."
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)
```

To visualise the shrinkage, the observed machine means are compared with
their posterior means. The line joining the two values for each machine
shows the direction and magnitude of shrinkage. The dashed horizontal
line represents the posterior mean of the population-level parameter
\(\mu\).

```{r shrinkage-plot, echo=FALSE, fig.cap="Shrinkage of the observed machine means towards the posterior population mean.", fig.pos="H", fig.align="center", fig.width=7, fig.height=4.5}
mu_posterior_mean <- mean(mu_draws)

plot(
  machine_levels,
  ybar_j,
  pch = 16,
  ylim = range(c(
    ybar_j,
    theta_posterior_means,
    mu_posterior_mean
  )),
  xlab = "Machine",
  ylab = "Mean deviation from 500 ml (ml)",
  main = "",
  xaxt = "n"
)

axis(
  1,
  at = machine_levels,
  labels = machine_levels
)

segments(
  x0 = machine_levels,
  y0 = ybar_j,
  x1 = machine_levels,
  y1 = theta_posterior_means
)

points(
  machine_levels,
  theta_posterior_means,
  pch = 17
)

abline(
  h = mu_posterior_mean,
  lty = 2,
  lwd = 1.2
)

legend(
  "topright",
  legend = c(
    "Observed machine mean",
    "Posterior machine mean",
    "Posterior population mean"
  ),
  pch = c(16, 17, NA),
  lty = c(NA, NA, 2),
  bty = "n"
)
```

The posterior means demonstrate the shrinkage expected under the
hierarchical model. Each machine-specific posterior mean is pulled from
its observed sample mean \(\overline{y}_j\) towards the common population
mean. For example, Machine 2 has an observed mean deviation of
$4.640$ ml, but its posterior estimate is pulled noticeably downwards
towards the population mean. Machine 1 also experiences visible
shrinkage towards the population mean.

The amount of shrinkage decreases as the number of observations for a
machine increases. Machine 1 has only \(n_1=5\) observations and therefore
borrows relatively more information from the population. In contrast,
Machine 7 has \(n_7=50\) observations, so its observed sample mean is more
precise and its posterior mean remains much closer to its observed mean.

This agrees with the result from part (b),

$$
m_j
=
w_j\overline{y}_j
+
(1-w_j)\mu,
$$

where

$$
w_j
=
\frac{\tau^2}
{\tau^2+\sigma^2/n_j}.
$$

As \(n_j\)increases, \(w_j\) approaches one, placing greater weight on the
machine-specific sample mean and producing less shrinkage. Thus the
hierarchical model performs partial pooling: machines with less data
borrow more information from the overall population, while machines with
more data are influenced more strongly by their own observations.

# Question 5

### (a)

The distribution of $Y_{i,\text{new}}\mid\mu$ is obtained by integrating
$\theta_{\text{new}}$ out of the joint density formed from the
observation model $Y_{i,\text{new}}\mid\theta_{\text{new}}$ and the
prior $\theta_{\text{new}}\mid\mu$:

$$
f(y_{i,\text{new}}\mid\mu)
=
\int_{-\infty}^{\infty}
f(y_{i,\text{new}}\mid\theta_{\text{new}})\,
\pi(\theta_{\text{new}}\mid\mu)\,
d\theta_{\text{new}},
$$

where

$$
Y_{i,\text{new}}\mid\theta_{\text{new}}\sim N(\theta_{\text{new}},\sigma^2),
\qquad
\theta_{\text{new}}\mid\mu\sim N(\mu,\tau^2).
$$

Since \(Y_{i,\mathrm{new}}=\mu+u_{\mathrm{new}}+\varepsilon_{i,\mathrm{new}}\), where \(u_{\mathrm{new}}\sim N(0,\tau^2)\) and \(\varepsilon_{i,\mathrm{new}}\sim N(0,\sigma^2)\) independently, it is a sum/convolution of independent Normal variables.Hence, the
mean and variance of $Y_{i,\text{new}}\mid\mu$ can be obtained using the
tower property and the law of total variance, conditioning on
$\theta_{\text{new}}$.

\textbf{Mean, by the tower property:}

$$
\begin{aligned}
\operatorname{E}(Y_{i,\text{new}}\mid\mu)
&=\operatorname{E}\Big[\operatorname{E}\big(Y_{i,\text{new}}\mid\theta_{\text{new}}\big)\,\Big|\,\mu\Big]\\
&=\operatorname{E}\big[\theta_{\text{new}}\mid\mu\big]\\
&=\mu.
\end{aligned}
$$

\textbf{Variance, by the law of total variance:}

$$
\begin{aligned}
\operatorname{Var}(Y_{i,\text{new}}\mid\mu)
&=\operatorname{E}\Big[\operatorname{Var}\big(Y_{i,\text{new}}\mid\theta_{\text{new}}\big)\,\Big|\,\mu\Big]
  +\operatorname{Var}\Big[\operatorname{E}\big(Y_{i,\text{new}}\mid\theta_{\text{new}}\big)\,\Big|\,\mu\Big]\\
&=\operatorname{E}\big[\sigma^2\mid\mu\big]
  +\operatorname{Var}\big[\theta_{\text{new}}\mid\mu\big]\\
&=\sigma^2+\tau^2.
\end{aligned}
$$

Therefore the distribution of $Y_{i,\text{new}}\mid\mu$ is:

$$
\boxed{
Y_{i,\text{new}}\mid\mu
\sim
N\left(\mu,\ \sigma^2+\tau^2\right).
}
$$

### (b)

We want to find \(\operatorname{E}[Y_{i,\text{new}}\mid\mathbf{y}]\) so by
the tower property. The full posterior predictive distribution is given by

\[
p(y_{i,\text{new}}\mid\mathbf{y})
=
\int
\phi(y_{i,\text{new}};\mu,\sigma^2+\tau^2)\,
\pi(\mu\mid\mathbf{y})\,d\mu,
\]

where \(\phi(\cdot\,;\mu,\sigma^2+\tau^2)\) denotes the Normal density with
mean \(\mu\) and variance \(\sigma^2+\tau^2\), from part (a). Rather than
evaluating this integral directly, the mean is found using the tower
property:

\[
\operatorname{E}[Y_{i,\text{new}}\mid\mathbf{y}]
=
\operatorname{E}_\mu\Big[\operatorname{E}(Y_{i,\text{new}}\mid\mu)\,\Big|\,\mathbf{y}\Big]
=
\operatorname{E}[\mu\mid\mathbf{y}].
\]

From part (a) we know \(\operatorname{E}(Y_{i,\text{new}}\mid\mu)=\mu\),
hence

\[
\operatorname{E}[Y_{i,\text{new}}\mid\mathbf{y}]
=
\operatorname{E}[\mu\mid\mathbf{y}]
\]

as having no data the mean relies entirely on the population mean, which
is the same as Question 3(d)(i):

\[
\boxed{
\widehat{\operatorname{E}[Y_{i,\text{new}}\mid\mathbf{y}]}
=
\widehat{\operatorname{E}[\mu\mid\mathbf{y}]}
=
0.373 \text{ ml}.
}
\]

```{r, echo = FALSE, fig.cap="Posterior predictive density of the deviation of a single bottle from a brand-new, as-yet-unobserved machine, obtained by simulating over the posterior draws of $\\mu$.", fig.pos="H", fig.align="center", fig.width=7, fig.height=4.2}
set.seed(69)  

y_new_draws <- rnorm(
  length(mu_draws),
  mean = mu_draws,
  sd = sqrt(sigma2 + tau2)
)

posterior_predictive_mean <- mean(mu_draws)

plot(
  density(y_new_draws),
  lwd = 2,
  xlab = "Simulated deviation of a new machine's bottle from 500ml (ml)",
  ylab = "Posterior predictive density",
  main = "Posterior predictive distribution of a new bottle's deviation"
)
abline(v = 0, lty = 2, lwd = 1.2)

```

### (c)

```{r, echo=FALSE}
mu_hat <- sum(ybar_j / V_j) / sum(1 / V_j)

mu_hat_table <- data.frame(
  `Quantity` = "Empirical Bayes estimate of mu",
  `Estimate` = round(mu_hat, 3),
  check.names = FALSE
)

mu_hat_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c"),
    caption = "Empirical Bayes point estimate of the population mean."
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)

full_predictive_mean <- mean(mu_draws)
full_predictive_variance <- sigma2 + tau2 + var(mu_draws)

eb_predictive_mean <- mu_hat
eb_predictive_variance <- sigma2 + tau2

comparison_table <- data.frame(
  `Quantity` = c(
    "Mean",
    "Variance"
  ),
  `Full posterior predictive` = c(
    round(full_predictive_mean, 3),
    round(full_predictive_variance, 3)
  ),
  `Empirical Bayes predictive` = c(
    round(eb_predictive_mean, 3),
    round(eb_predictive_variance, 3)
  ),
  check.names = FALSE
)

comparison_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c", "c"),
    caption = "Comparison of the full posterior predictive and empirical Bayes predictive means and variances."
  ) |>
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE)

set.seed(69)
y_new_eb_draws <- rnorm(length(mu_draws),mean = mu_hat,sd = sqrt(sigma2 + tau2))

plot(density(y_new_draws),lwd = 2,col = "blue",xlab = "Simulated deviation of a new machine's bottle from 500ml (ml)",ylab = "Density",main = "Posterior predictive density: full posterior vs. empirical Bayes",ylim = range(0, density(y_new_draws)$y, density(y_new_eb_draws)$y))

lines(density(y_new_eb_draws),lwd = 2,col = "red",lty = 2)

legend("topright", legend = c( expression(p(y[i*","*new] ~ "|" ~ y)), expression(p(y[i*","*new] ~ "|" ~ y*","*hat(u)))),col = c("blue", "red"),lwd = 2,lty = c(1, 2))

```


The full posterior predictive distribution has mean
\(\operatorname{E}(Y_{i,\text{new}}\mid\mathbf{y})\approx0.373\) ml and
variance \(\operatorname{Var}(Y_{i,\text{new}}\mid\mathbf{y})\approx6.580\),
whereas the empirical Bayes  distribution has mean
\(\hat\mu\approx0.420\) ml and variance \(6.25\). The full Bayesian
distribution is centered slightly closer to zero because it retains the
influence of the zero-centered Laplace prior. It is also slightly wider
because it additionally accounts for posterior uncertainty in \(\mu\),
whereas the empirical Bayes approach treats \(\hat\mu\) as known exactly.
Both methods already incorporate the machine-to-machine variance
\(\tau^2\) and bottle-to-bottle variance \(\sigma^2\); the only difference
between them is whether \(\operatorname{Var}(\mu\mid\mathbf{y})\) is
additionally included. The curves remain visually similar because this
extra term is small relative to \(\sigma^2+\tau^2\).
